# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
### Assumptions
1. Working directory is set to current local clone of the github repository for this assignment.
2. The dataset, `activity.zip`, required for the reproducible research is already downloaded to the repository. 

```{r loadData, echo=TRUE}

data.local.zip = "activity.zip"
data.local = "activity.csv"

# extract the file, if not already extracted 
if (!file.exists (data.local))  
    unzip (data.local.zip)

# fail if there is a problem with extracted data
if (!file.exists (data.local))          
    stop ("unable to extract the source data!") 

# load the data
activityData <- read.csv(data.local,na.strings="NA",stringsAsFactors=FALSE);

# take a quick look at the data culture
summary (activityData)  
#str (activityData)  
#head (activityData)  
#tail (activityData)  

```


## What is mean total number of steps taken per day?

- Make a histogram of the total number of steps taken each day
```{r histogram,fig.width=8, fig.height=6,echo=TRUE}
require(ggplot2);
require(plyr);
totalSteps <- ddply(activityData, .(date),function(activityData)sum(activityData$steps));
names(totalSteps) <- c("Date","Total Steps");
#summary (totalSteps);

# plot the histogram
hist(totalSteps[["Total Steps"]],main="Total no. steps for each days",xlab="Steps");


```


- Calculate and report the mean and median total number of steps taken per day
```{r mean,echo=TRUE}
meanValue <- mean(totalSteps[["Total Steps"]],na.rm=TRUE);
medianValue <- median(totalSteps[["Total Steps"]],na.rm=TRUE);

print(meanValue);
```
Mean : `r meanValue`      
Median: `r medianValue`



## What is the average daily activity pattern?

- Time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

plotGraph <- function(dataSet){
    # Initialize ggplot
    g<-ggplot(dataSet, aes(interval,steps));
    # Add layer
    g <- g+geom_line(size=1);
    # Add label
    g <- g+labs(title="Average Daily Activity Pattern", x="Interval", y="Steps");
    
    print(g);
    
    }
# Find the average steps for interval
avrStepPerInterval <- ddply(activityData,
                            .(interval),
                            function(activityData)mean(activityData$steps,na.rm=TRUE));

names(avrStepPerInterval) <- c("interval","steps");

plotGraph(avrStepPerInterval);
```


- 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
```{r}
# Find interval in which maximum steps contained
maxStepInterval <- avrStepPerInterval[which.max(avrStepPerInterval$steps),]$interval;

```
Interval contains the maximum number of steps: `r maxStepInterval`




## Imputing missing values


- Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r missingDataReport,echo=TRUE}
## Remove NA values
rowCount <- nrow(activityData[is.na(activityData$steps),]);

```
The total number of rows with NAs is `r rowCount`.


- Fill all of the missing values in the dataset with mean value of the interval
```{r healling dataset, echo=TRUE}
meansValue <- function(activityData, avrStepPerInterval) {
    naIndices <- which(is.na(activityData$steps));    
    replacements <- unlist(lapply(naIndices, FUN=function(x){
        interval = activityData[x,]$interval;        
        temp <- round(avrStepPerInterval[which(avrStepPerInterval$interval == interval),]$steps,2);        
        }))
    missedSteps <- activityData$steps;   
    missedSteps[naIndices] <- replacements;
    missedSteps;
    }
## Data frame for fixed data
healedData <- data.frame(
    steps = meansValue(activityData, avrStepPerInterval),
    date = activityData$date,
    interval = activityData$interval);

origSummary <- summary(activityData);
filledDataSummary <- summary(healedData);
```
Summary of original data: `r origSummary`

Summary of filled data: `r filledDataSummary`



- Histogram of the steps taken each days
```{r histogramForHealData, echo=TRUE}
steps <- ddply(healedData,.(date),function(healedData)sum(healedData$steps));
names(steps) <- c("Date","Total Steps");
hist(steps[["Total Steps"]],main="Total no. steps for each days",xlab="Steps");

mean <- round(mean(steps[["Total Steps"]],na.rm=TRUE),2);
median <- round(median(steps[["Total Steps"]],na.rm=TRUE),2);

```
Mean after fixing NA values : `r mean`      
Median after fixing NA values: `r median`



## Are there differences in activity patterns between weekdays and weekends?

```{r weeklyActivity, echo=TRUE}
evaluateWeeklyStep <- function(fixedData) {
    fixedData$weekday <- as.factor(weekdays(as.Date(fixedData$date)));
    weekend <- subset(fixedData, weekday %in% c("Saturday","Sunday"));
    weekday <- subset(fixedData, !weekday %in% c("Saturday","Sunday"));
    
    weekendSteps <- ddply(weekend,.(interval),function(weekend)mean(weekend$steps,na.rm=TRUE));    
    weekdaySteps <- ddply(weekday,.(interval),function(weekday)mean(weekday$steps,na.rm=TRUE));
    
    weekendSteps$dayofweek <- rep("weekend", nrow(weekendSteps));
    weekdaySteps$dayofweek <- rep("weekday", nrow(weekdaySteps));
    
    week <- rbind(weekendSteps, weekdaySteps);
    week$dayofweek <- as.factor(week$dayofweek);
    names(week) <- c("interval","steps","dayofweek");
    week;
    }
weeklyPlot <- function(healedData) {
    ggplot(healedData,
           aes(interval,steps)) +
        geom_line(size=1) +
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Steps");        
    }
weekData <- evaluateWeeklyStep(healedData);
weeklyPlot(weekData);
```
